---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r, echo=TRUE}
# Unzip file and read in resulting csv
unzip('activity.zip')
activity <- read.csv('activity.csv')

# date read in as factor--convert to Date format
activity$date <- as.Date(activity$date)
```


## What is mean total number of steps taken per day?
```{r, echo=TRUE}
# Ignore missing values
act.no.na <- activity[!is.na(activity$steps),]

# Make histogram of total number of steps taken each day
steps.per.day <- aggregate(act.no.na$steps, by = list(act.no.na$date), sum, 
                           na.rm = TRUE)
hist(steps.per.day$x, xlab = "Total steps per day", ylab = "Number of days", 
     main = "Distribution of total steps taken each day" )

# Do this so that mean displayed below won't be in scientific notation
options(scipen = 999)
```

The mean total number of steps taken per day was `r mean(steps.per.day$x)` and the median total number of steps taken per day is `r median(steps.per.day$x)`.

## What is the average daily activity pattern?

```{r, echo=FALSE}
# Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) 
# and the average number of steps taken, averaged across all days (y-axis)
steps.per.int <- aggregate(activity$steps, by = list(activity$interval),
                           mean, na.rm = TRUE)
plot(x = steps.per.int$Group.1, y = steps.per.int$x, type = "l",
     xlab = "5-minute nterval", ylab = "Average steps",
     main = "Average number of steps taken per interval")
```

The five-minute interval containing the maximum number of steps on average across all days in the dataset is the one starting at `r steps.per.int$Group.1[which.max(steps.per.int$x)]` minutes.

## Imputing missing values
```{r, echo = FALSE}
# Calculate and report the total number of missing values in the dataset (i.e. 
# the total number of rows with NAs)

# Devise a strategy for filling in all of the missing values in the dataset. 
# The strategy does not need to be sophisticated. For example, you could use 
# the mean/median for that day, or the mean for that 5-minute interval, etc.

# Create a new dataset that is equal to the original dataset but with the 
# missing data filled in.

# Make a histogram of the total number of steps taken each day and Calculate 
# and report the mean and median total number of steps taken per day. Do these 
# values differ from the estimates from the first part of the assignment? What 
# is the impact of imputing missing data on the estimates of the total daily 
# number of steps?
```


## Are there differences in activity patterns between weekdays and weekends?
